---
title: "The Data Druid's Garden"
subtitle: "On Biodiversity, Life, and Growth"
author: "Jake Weber"
date: "January 24, 2022"
output: 
  html_document:
    theme: yeti
    highlight: kate
    code_folding: show
    toc: TRUE
    toc_depth: 2
    number_sections: true
    toc_float:
      collapsed: false
      smooth_scroll: true
editor_options: 
  chunk_output_type: inline
  markdown: 
    wrap: 72
---

```{r INITIATING, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Hello to all who may stumble upon this. Here is my exploration into a
plant inventory analysis. The data embodies my personal plant collection
so it was especially fun for me to actually quantify and analyze them.
Lastly, you'll find a visualization at the end of this called **Life in
a Lock** which is my attempt to blend a data-driven analysis with some
creative manipulation that I call art. Hope you enjoy!

For those interested, please feel free to get in contact via:

-   LinkedIn: [Jake Weber](https://www.linkedin.com/in/mrjakeweber/)

-   Github: [Jake Weber](https://github.com/witnwhimsy)

-   Email: Mr.Jake.Weber\@gmail.com

# Preparation

## Packages

The **tidyverse** is the main workhorse of my R packages. Almost
everything vital stems from it's ecosystem and supplements the suite.

```{r message=FALSE, warning=FALSE}
#options(ggplot2.continuous.color = "viridis")
#options(ggplot2.continuous.fill = "viridis")
#options(ggplot2.discrete.fill = "viridis")

options(scipen = 999)
setwd("C:/Users/mrjak/Desktop/Data Vizardry/Data Druid")

#packages ----

#Workhorse
library(tidyverse)

#Import & Export
library(readxl)
library(writexl)
library(readr)

#Formatting & Visualization
library(tidyquant)
library(hrbrthemes)
library(kableExtra)
library(viridisLite)
library(scales)
library(DT)
library(ggrepel)
```

## Reading the Data

Ingesting the data from a .csv that I originally built in Google Drive
which could lend it's self to automated, on-going reporting.

```{r warning=FALSE}
import_dat <- readr::read_csv("Arcadian Gardens Inventory Analysis.csv")
```

## Glimpse of the Data

Using **glimpse()** to take a peak at the data with an easy-to-read
summary.

```{r}
import_dat %>% glimpse()
```

# Data Manipulation

## Renaming the Columns

Altering the names of the columns to make them more manageable in R. We
can always create prettier labels later on.

```{r}
dat <- import_dat %>%
    rename( #renaming the columns
        id = id,
        name_scientific = `Scientific Name`,
        name_common = `Common Name`,
        maturity = Maturity,
        water = `Watering Preference`)
```

## Fixing a Typo

Using a quick summary, we can see that one of the "adults" is missing a
letter. Then, we account for this via **case_when()** to patch the data.
This fixes it for any downstream inquiries into Maturity.

### Spotting the Typo

```{r}
dat %>% group_by(maturity) %>% summarise(n = n()) %>% 
    kbl(align = "l", format.args = list(big.mark = ",")) %>% 
    kable_styling(
        full_width = F,
        bootstrap_options = c("hover", "responsive", "striped"))
```

### Fix with Case When

```{r}
dat <- dat %>% 
    mutate( #tidying text and a typo in the data
        name_scientific = name_scientific %>% str_to_title(),
        maturity = case_when(maturity == "Adut" ~ "Adult",
                             T ~ maturity),
        name_scientific = case_when(name_scientific == "Saintpaulia" ~ "Saintpaulia Incognita",
                                    T ~ name_scientific))
```

## Deriving Genus x Species

We are given the Genus and Species via the Scientific Name, the trick is
extracting them. First, I use **separate()** to break out the *genus*
since they are consistently one word in my collectiona and followed by a
space. <br> <br> The species were a little trickier due to having a
variety of patterns that broke separate(), such as 2+ word species,
apostrophization, hybrids & cultivars. So for the second part, I simply
subtracted the new *genus* from the existing Scientific name via
**str_remove()** to derive the *species*.

### Engineering Genus

```{r}
dat <- dat %>% 
    separate(col = name_scientific, 
             into = c("genus"), 
             sep = " ", 
             remove = FALSE) %>% #pulling the genus out of the scientific genus species
    select(genus, contains("name"), water, maturity, id) %>% 
    mutate(genus = genus %>% as_factor(),
           name_common = name_common %>% str_to_title()) 

dat %>% distinct(genus) %>% pull()
```

### Engineering Species

```{r}
dat <- dat %>% 
    mutate(species = str_remove(as.character(name_scientific), 
                                as.character(genus)) %>% 
               trimws("both")) %>% #using genus to derive species from scientific name
    select(genus, species, contains("name"), maturity, water, id)

dat %>% distinct(species) %>% pull()
```

## Tidy'd Data

```{r}

dat %>% datatable(filter = "bottom")

```

## Stat'd Data

```{r}

dat2 <- dat %>%  # adding some stats
    group_by(genus) %>% mutate(n_genus = n()) %>% ungroup() %>% 
    group_by(name_scientific) %>% mutate(n_species = n()) %>% ungroup()  %>% 
    arrange(desc(n_genus), desc(n_species)) %>% 
    mutate(genus = as_factor(genus),
           species = as_factor(species),
           # genus_scaled = n_genus / max(n_genus) * 100,
           # species_scaled = n_species / max(n_species) * 100,
           perc_genus = n_genus / sum(n_genus) * 100,
           perc_species = n_species / sum(n_species) * 100,
           percent_genus = scales::percent(n_genus / sum(n_genus)),
           percent_species = scales::percent(n_species / sum(n_species))) %>% 
    select(name_scientific, genus, species, contains("n_"), contains("perc"), everything())

dat3 <- dat2 %>%  # creating a summary of species collected
    group_by(genus) %>% 
    mutate(test = paste0(species)) %>% 
    summarise(genera_collected = toString(unique(species))) %>% 
    ungroup() 

plant_dat <- left_join(x = dat2, y = dat3, by = c("genus")) %>% select(1,2,3,genera_collected,everything())

plant_dat %>% 
    select(genus, name_scientific, name_common, genera_collected, 
           contains("n_"), contains("percent_")) %>% 
    datatable(filter="bottom" )

```

# Exploration

## Basic Overview

We see here that I have 73 plants in my collection which represents 22
unique genera and 49 unique species.

```{r}

plant_dat %>% 
    summarise(
        n = n(),
        n_distinct_genus = n_distinct(genus),
        n_distinct_species = n_distinct(name_scientific)
    ) %>% 
    kbl(align = "l", format.args = list(big.mark = ",")) %>% 
    kable_styling(
        full_width = F,
        bootstrap_options = c("hover", "responsive", "striped"))

```

## Genera

### Summary

```{r}
viz_dat_genus <- 
    plant_dat %>% 
    group_by(genus, genera_collected) %>% summarise(n = n()) %>% ungroup() %>% 
    mutate(percent_total = scales::percent(n/sum(n)))

viz_dat_genus %>% kbl(align = "l", format.args = list(big.mark = ",")) %>% 
    kable_styling(
        full_width = F,
        bootstrap_options = c("hover", "responsive", "striped"))
```

### Viz

```{r fig.height=6, fig.width=9}

viz_genus <- 
    viz_dat_genus %>% 
    ggplot(aes(text = str_glue(
    "Genera: {genus},
     Count: {n},
     Percent of Collection: {percent_total}
     Species Collected: {genera_collected}"))) +
    geom_col(aes(x = genus %>% fct_reorder(n), fill = n,y = n)) +
    scale_y_continuous(breaks = seq(0, 12.5, by = 3)) +
    coord_flip() +
    scale_fill_viridis_c(option = "viridis", direction = 1) +
    theme_minimal() +
    labs(
        title = "Favorite Genera",
        subtitle = "Can you tell I like Anthuriums?",
        caption = "Out of the 22 Genera represented, my top 5 represent 47.5% of my collection",
        x = "Genera",
        y = "Count",
        fill = "Count")

viz_genus
```

### Interactive Viz

```{r fig.height=6, fig.width=9}
viz_genus %>% plotly::ggplotly(tooltip = "text")
```

## Genus Species

Next, I'm going to visualize the species found with the various genera.

### Top 5 Genera

To start, I'm honing in on the top 5 genera within my collection. The
generan selected are: Anthurium, Ficus, Monstera, Begonia, &
Philodendron. <br> <br> I call this visualization **Life in a Lock** as
a saw a parallel between the visualization and that of a keyhole. "Life"
stemming from how the visualization depicts biodiversity within my
collection.

```{r fig.height=9, fig.width=9}
# library
library(tidyverse)
 
# Create dataset
data <- plant_dat %>% filter(n_genus >= 5) %>% 
    group_by(genus, name_scientific, species) %>% 
    summarise(n = n()) %>% ungroup()

  data$genus <- factor(data$genus, 
         levels = c("Anthurium", 
                    "Ficus",
                    "Monstera",
                    "Begonia",
                    "Philodendron")) 
  
   data$species <- factor(data$species,
                levels = c("Clarinervium","Crystallinum",
                           "Magnificum","Regale",
                           "Dorayaki","Regale+Magnificum", 
                           "Lyrata","Elastica 'Ruby'" ,
                           "Benghalensis","Elastica 'Tineke'",
                           "Elastica 'Burgundy'",
                           "Deliciosa 'Albo Variegata'",
                           "Deliciosa",
                           "Adansonii","Maculata", 
                           "'Lois Burke'","Fibrous 'My Special Angel'",
                           "Fibrous 'Whimsy'","Verrucosum 'Mini'",
                           "'El Choco'","Erubescens",
                           "Micans","Verrucosum X Melanochrysum",
                           "Verrucosum"))
      
data <- data %>% 
      mutate(group = genus,
             individual = species,
             value = n) %>% 
      select(group, individual, value)



# Set a number of 'empty bar' to add at the end of each group
empty_bar <- 4
to_add <- data.frame( matrix(NA, empty_bar*nlevels(data$group), ncol(data)) )
colnames(to_add) <- colnames(data)
to_add$group <- rep(levels(data$group), each=empty_bar)
data <- rbind(data, to_add)
data <- data %>% arrange(group)
data$id <- seq(1, nrow(data))
# 
# label_data <- data
# number_of_bar <- nrow(label_data)
# label_data$angle <- 360 * (label_data$id-0.5)/number_of_bar     # I substract 0.5 because the letter must have the angle of the center of the bars. Not extreme right(1) or extreme left (0)
# # label_data$hjust <- ifelse(angle < -90, 1, 0)
# # label_data$angle <- ifelse(angle < -90, angle+180, angle)
 
# Make the plot
p <- ggplot(data, aes(x=individual, y=value, fill = group)) +  
  geom_bar(stat="identity") +
  ylim(-1,5) +
  scale_fill_viridis_d(option = "viridis", direction = 1) +
  theme_minimal() +
    labs(
        title = "Life in a Lock",
        subtitle = "Biodiversity amongst my top Genera",
        caption = "Out of the 22 Genera represented, my top 5 represent 47.5% of my collection",
        x = "",
        y = "Count",
        fill = "Genera") +
    theme(
      legend.position = "bottom",
#     axis.text = element_blank(),
#     axis.text.x = element_text(angle = 90),
#     axis.title = element_blank(),
#      panel.grid = element_blank(),
     plot.margin = margin(.5, .5, .5, .5, "cm")
)  +
  coord_polar(start = 3.05, clip = "off")
    

 
p

```

```{r}

ggsave(plot = p, filename = "Life in a Lock_minimal.pdf", width = 9, height = 9, units = "in")


```
